Hippocampal replay of experience at real world speeds

### 2021-05-13 Article.tex

\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The rat hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at \texttildelow20x movement speed of the animal. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. Using this model, we find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \citep{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience.

Hippocampal "replay" events are well suited to that role \citep{CarrHippocampalreplayawake2011}. As rats move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and are thus thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow5-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \citep{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009, Tingleymethodsreactivationreplay2020}. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent some definition of a "random" sequence.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test. This is problematic because it has the potential to mischaracterize real events that do not move at constant speeds such as those that change direction or are discontinuous \citep{LiuMethodsAssessmentMemory2018}. Furthermore, the use of large, fixed-size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. For example, with 20 ms time bins and 3 cm position bins, an event can only move in 1.5 m/s increments between time steps and all spikes within the 20 ms bins are combined. The linear fit is also highly dependent on the estimation of event boundaries, such as the start and end times of a SWR, because the fit depends on all the data over the course of the event. Approaches that focus on the order of cell activity within each event \citep{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches often exclude events that have stationary representations of a single location \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \citep{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

We find that the large majority of SWRs contain spatially coherent content; that is, trajectories that are spatially concentrated at each moment in time and have no large discontinuities in position. Surprisingly, while the expected high-speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

\subsection*{Overview of the model}
We begin with an application of the state space model to simulated data, both to validate the model and to provide intuition (\FIG{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence (\FIG{1}A) and apply our model to the sequence. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely). Last, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

@@

This shows off two key features of the model. First, the fragmented dynamic gives us a way to directly test for spatially incoherent content for each time, rather than by the use of non-parametric shuffle, as is commonly used with the standard decoder, because we have a model for spatial incoherent content. Second, this approach allows the model to to capture a wide range of movement speeds for the latent position in contrast to the standard decoder line fit (red dashed line in \FIG{1}E). The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (\FIG{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to each of our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), but there are also intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these intermediate speeds as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, \FIG{1}F) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. In summary, we can characterize the speed or set of speeds in a SWR based on the probability of each of the three movement dynamics over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

We next sought to validate the model on real hippocampal data recorded from awake, behaving rats. To ensure that we could capture rapid trajectories many times the speed of the animal---as described by most studies of hippocampal replay---we first decoded a single SWR with sequential population activity (\FIG{2}A, top panel) from 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track (see Methods for details). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary dynamic was also noticeable at the beginning and end of the SWR (\FIG{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of latent linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (\FIG{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D to better see the spatial trajectory on the maze (\FIG{2}C). Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (\FIGSUPP[2]{fig2sup1}A).

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019}, is time-consuming, and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sequences and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run, demonstrating the power of our algorithm. We also verified that the population spiking for all positions on the track was consistent for each animal (\FIGSUPP[2]{fig2sup3}).

Having established the validity of our decoder in tracking the animal's position, we next wanted to assess the overall prevalence of spatial content across SWRs. We detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (\FIGSUPP[2]{fig2sup2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,071 of 25,844) of SWRs as containing at least one of the three dynamics or dynamic mixtures. To ensure that this depends on the firing of spatially tuned cells in the hippocampus, we resample position with replacement, which shuffles the relationship between spiking and position, for two recording sessions, fit the encoding model and then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}). This shows that our model will not classify dynamics when there is no spatial information present in the data.

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffle that randomized the order of the reward well to reward well runs and then circularly permuted the run. This, in contrast to our previous shuffle, should preserve correlation between spikes locally but destroy global spatial structure. Using this shuffle, we found that there was more shuffled spatially incoherent content compared to the real data and there was less spatially coherent content compared to the real data for two recording sessions (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup3}).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4,449 of 23,071 or 19\% of classified SWRs had time periods where the decoded position was classified as continuous (\FIG{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories excludes a large fraction of events where there is evidence for spatially coherent, but not continuously changing, content. Note that here we did not limit our analysis to only those SWRs that had continuous dynamics for the entire SWR, as is assumed by the standard decoder in the case of fitting a line, because our decoder is much more flexible in terms of allowing different speeds during the SWR event.

@@

For the linear regression method, we draw 1000 position samples from each time bin according to the probability of position in that time bin. A linear regression is then fit to the sampled positions using time as a covariate. Fits were made separately for the posterior probabilities corresponding to the center arm with the left arm and the center arm with the right arm. Quality of the fit was assessed using the R-squared and the best fit between the arms was selected based on this metric.

>>>>

\begin{abstract}
Representations related to past experiences play a critical role in memory and decision-making processes. The rat hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at \texttildelow20x movement speed of the animal. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. Using this model, we find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, many events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}

@@

The brain has the remarkable ability to store and retrieve representations of past events, allowing memories of the past to influence the future. These memory processes depend critically on the hippocampus, where rapid plasticity during an experience is thought to drive the initial encoding of events (Cite Eichenbaum and Cohen?). Subsequently, hippocampal "replay" of stored representations during both slow-wave sleep and periods of immobility during waking contributes to the longer term storage and updating of event memories in distributed hippocampal-neocortical circuits (REFs). Experiences can take place over many seconds, minutes, hours or days, and to be useful in guiding behavior, memories of these events must be stored in such a way that they can be retrieved quickly to inform upcoming choices. Consistent with this need for rapid storage, hippocampal replay events can express time compressed representations of past experience, activating regions outside the hippocampus and presumably driving plasticity in those regions to promote memory storage. \citep{CarrHippocampalreplayawake2011}.

In the context a rodent navigating through space preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and thus "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow5-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \citep{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009, Tingleymethodsreactivationreplay2020}. One might therefore conclude that all such events are sequential in nature, and thus that memories that are stored or updated by SWRs are those corresponding sequences of locations. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance, however. Such memories would seem to be useful, in that they could encode the stimuli and values associated with a given place irrespective of the path used to get there. If the rodent hippocampus is capable of storing those sorts of memories, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a brief pattern  more similar to one seen during a real experience.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, although another report suggested that events that replay a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations or a small region of the environment, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approaches for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent an implicit definition of a "random" sequence.

While the standard approach identifies constant speed events, it does not consider events that are rejected by the statistical test. This is problematic because it has the potential to mischaracterize real events that do not move at constant speeds such as those that change direction or are discontinuous \citep{LiuMethodsAssessmentMemory2018}. Furthermore, the use of large, fixed-size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. For example, with 20 ms time bins and 3 cm position bins, an event can only move in 1.5 m/s increments (1 or more bins in 20 ms) between time steps. The linear fit is also highly dependent on the estimation of event boundaries, such as the start and end times of the SWR, because the fit depends on all the data over the course of the event. Approaches that focus on the order of cell activity within each event \citep{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches often exclude events that have stationary representations of a single location \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

We therefore developed a state space model that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \citep{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

We find that the large majority of SWRs contain spatially coherent content; that is, trajectories that are spatially concentrated at each moment in time and have no large discontinuities in position. Surprisingly, while the expected high-speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real-world experiences. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

\subsection*{Overview of the model}
We begin with an application of the state space model to simulated data, both to validate the model and to provide intuition (\FIG{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, and fields are spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence (\FIG{1}A) and apply our model to the sequence. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (at the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely). Last, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely. Importantly, the fragmented dynamic provides a well-specified definition of what we mean by random, or unstructured activity.

@@

This illustrates two key features of the model. First, as mentioned above, the fragmented dynamic gives us a way to directly test for spatially incoherent content for each time, rather than by the use of non-parametric shuffle, as is commonly used with the standard decoder. Second, this approach allows the model to to capture a wide range of movement speeds for the latent position in contrast to the standard decoder line fit (red dashed line in \FIG{1}E). The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (\FIG{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to each of our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), but there are also intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these intermediate speeds as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, \FIG{1}F) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. In summary, we can characterize the speed or set of speeds in a SWR based on the probability of each of the three movement dynamics over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

We next sought to validate the model on real hippocampal data recorded from awake, behaving rats. To ensure that we could capture rapid trajectories many times the speed of the animal---as described by most studies of hippocampal replay---we first decoded a single SWR with sequential population activity (\FIG{2}A, top panel). Here the rat was performing a spatial alternation task on a W-shaped track (see Methods for details). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary dynamic was also noticeable at the beginning and end of the SWR (\FIG{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay is initially slower--- a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of latent linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (\FIG{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D to better see the spatial trajectory on the maze (\FIG{2}C). Importantly, when we apply the same model using the 2D position of the animal we get a similar result (\FIGSUPP[2]{fig2sup1}A).

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019} and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sequences and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference between actual and decoded location is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run, demonstrating the power of our algorithm. We also verified that the population spiking for all positions on the track was consistent for each animal (\FIGSUPP[2]{fig2sup3}).

Having established the validity of our decoder in tracking the animal's position, we next assessed the prevalence of spatial content across SWRs. We detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observed replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using standard approaches (\FIGSUPP[2]{fig2sup2}A-F). However, we also observed many events with spatially coherent content that do not have this structure. For example, there are trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observed SWRs where the content is spatially incoherent throughout(\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, 89\% (23,071 of 25,844) of SWRs contain at least one of the three dynamics or dynamic mixtures. To ensure that this depends on the firing of spatially tuned cells in the hippocampus, we resampled position with replacement, which shuffles the relationship between spiking and position, for two recording sessions. We then fit the encoding model and then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}). This shows that our model does not identify dynamics when there is no spatial information present in the data.

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffle that randomized the order of the reward well to reward well runs and then circularly permuted the resulting data. This, in contrast to our previous shuffle, will preserve the local correlations between spikes but reduce global spatial structure. Using this shuffle, we found that there was still less spatially coherent and more spatially incoherent content compared to the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup3}).

To more directly compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4,449 of 23,071 or 19\% of classified SWRs had time periods where the decoded position was classified as continuous (\FIG{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories excludes a large fraction of events where there is evidence for spatially coherent, but not continuously changing, content. Note that here we did not limit our analysis to only those SWRs that had continuous dynamics for the entire SWR, as is assumed by the standard decoder in the case of fitting a line, because our decoder is much more flexible in terms of allowing different speeds during the SWR event.

@@

For the linear regression method, we drew 1000 position samples from each time bin according to the probability of position in that time bin. A linear regression was then fit to the sampled positions using time as a covariate. Fits were made separately for the posterior probabilities corresponding to the center arm with the left arm and the center arm with the right arm. Quality of the fit was assessed using the R-squared and the best fit between the arms was selected based on this metric.

For the MAP estimate, we pick the position bin with the maximum value for each time bin, which corresponds to the most probable position.
